# Call SAS program

Sometimes it's just useful to do some heavy lifting in SAS.
For example, using `proc sql` to join multiple tables simultaneously is easier to do in SAS.
(I know, there are some SQL packages for R)

The basic process for call SAS from R is

1. Write data to be processed by SAS as CSV.
2. Call an existing SAS program. **TO DO**
3. Read data file created by the SAS program. **TO DO**


## Create a fake dataset

Create a fake dataset for illustration.

```{r, tidy=FALSE}
s <- 1000
D <- data.frame(id = seq(1, s),
	            date = sample(seq(as.Date("2013-01-01"), as.Date("2013-12-31"), 1), s, replace=TRUE),
	            state = sample(state.name, s, replace=TRUE),
	            x1 = rnorm(s),
	            x2 = runif(s),
	            x3 = rbinom(s, 1, 0.5))
```

Recode some random values of x1 to `NA`.

```{r}
sample <- sample(D$id, s * 0.10)
D$x1[sample] <- NA
```

Examine the dataset.

```{r}
summary(D)
```


## Write data to CSV

Some things to note about the call to `write.csv`.

* `quote=FALSE` prevents writing redundant quotation marks
* `na="."` changes the R default `NA` to the SAS default `.` for missing values
* `row.names=FALSE` also prevents writing redundant stuff

```{r}
path <- getwd()
f <- file.path(path, "fakedata.csv")
write.csv(D, f, quote=FALSE, na=".", row.names=FALSE)
```

Read a few lines from the CSV file just to take a peek.

```{r}
readLines(f, n=5)
```
