# Call SAS program

Sometimes it's just useful to do some heavy lifting in SAS.
For example, using `proc sql` to join multiple tables simultaneously is easier to do in SAS.
(I know, there are some SQL packages for R)

The basic process for call SAS from R is

1. Write data to be processed by SAS as CSV.
2. Call an existing SAS program.
3. Read data file created by the SAS program.


## Create a fake dataset

Create a fake dataset for illustration.

```{r, tidy=FALSE}
s <- 1000
D <- data.frame(id = seq(1, s),
	            date = sample(seq(as.Date("2013-01-01"), as.Date("2013-12-31"), 1), s, replace=TRUE),
	            state = sample(state.name, s, replace=TRUE),
	            x1 = rnorm(s),
	            x2 = runif(s),
	            x3 = rbinom(s, 1, 0.5))
```

Recode some random values of x1 to `NA`.

```{r}
sample <- sample(D$id, s * 0.10)
D$x1[sample] <- NA
```

Examine the dataset.

```{r}
summary(D)
```


## Write data to CSV

Some things to note about the call to `write.csv`.

* `quote=FALSE` prevents writing redundant quotation marks.
* `na="."` changes the R default `NA` to the SAS default `.` for missing values; it will prevent an *Invalid data* note to be written to the log.
* `row.names=FALSE` also prevents writing redundant stuff.

```{r}
path <- getwd()
f <- file.path(path, "fakedata.csv")
write.csv(D, f, quote=FALSE, na=".", row.names=FALSE)
```

Read a few lines from the CSV file just to take a peek.

```{r}
readLines(f, n=10)
```


## Call SAS

Call SAS to

1. Read data.
2. Create new variable `x4`.

Define a function to make a SAS command string.
The command string writes a SAS log using the same file name as the program file, `f`.

```{r}
makeCmd <- function(f) {
  path <- getwd()
  sasFile <- file.path(path, paste0(f, ".sas"))
  logFile <- file.path(path, paste0(f, ".log"))
  sasexe <- file.path("C:", "Program Files", "SASHome2", "SASFoundation", "9.4", "sas.exe")
  str <- paste(
    paste0("\"", sasexe, "\""),
    "-sysin", paste0("\"", sasFile, "\""),
    "-log", paste0("\"", logFile, "\""),
    "-print", paste0("\"", logFile, "\"")
 )
  show(str)
  str
}
```

Use the SAS command string in a system call.
The SAS program file is `createNewVariable.sas`.
The program file was not automatically generated.
The SAS log file is `createNewVariable.log`.
**If there were any errors, then there will be an error code returned here.**
Check the SAS log for details on any errors.

```{r}
cmd <- makeCmd("createNewVariable")
system(cmd, invisible = FALSE)
```


## Read data file created by the SAS program

```{r}
path <- getwd()
f <- file.path(path, "fakedata.csv")
D <- read.csv(f)
```

Need to change the class of the date variable.
It is read in by `read.csv` as a character class.

```{r}
D$date <- as.Date(D$date)
```


Examine the dataset.

```{r}
summary(D)
```


## Appendix: the SAS program `createNewVariable.sas`

```{r}
path <- getwd()
sasFile <- file.path(path, "createNewVariable.sas")
readLines(sasFile)
```