---
title: "Alternatives to displaying protected health information"
author: "Benjamin Chan (chanb@ohsu.edu)"
date: "Wednesday, December 10, 2014"
output: html_document
---

# Context

Often, it is useful to document the structure of datasets in output so the reader can follow the flow of the code.
However, the datasets we use contain [protected health information (PHI)](http://www.hipaa.com/2009/09/hipaa-protected-health-information-what-does-phi-include/).
To allow for the ability to share code and output to external entities while maintaining HIPAA compliance, the following methods are proposed.

1. Hashing the PHI
2. Writing PHI to external datasets


# Read example dataset

**This dataset does not contain actual PHI.**

Source scripts to read in the CMS [Open Payments](http://www.cms.gov/OpenPayments) data.
Scripts are from the following [GitHub repository](https://github.com/benjamin-chan/OpenPayments).

```{r}
setInternet2()
url <- "https://raw.githubusercontent.com/benjamin-chan/OpenPayments/master/library.r"
source(url)
url <- "https://raw.githubusercontent.com/benjamin-chan/OpenPayments/master/getData.r"
source(url)
require(data.table)
```

For simplicity, operate on a subset of columns in data frame D.

```{r}
varToKeep <- c("General_Transaction_ID", "Physician_Profile_ID", "Physician_Last_Name", "Physician_Specialty")
D <- D[, varToKeep, with=FALSE]
```


# Example of non-compliance

Pretend all columns except `Physician_Specialty` are PHI.

The following output would **not** be HIPAA-compliant.

```{r}
head(D)
str(D)
```


# Example of hashing PHI

**Need to find a good hash/anonymizer/scrambler function**
For now, mask half of the characters in the string.

```{r}
mask <- function (s) {
  len <- nchar(s)
  s <- paste0("***", substr(s, len / 2, len))
  s
}
```

"Anonymize" the PHI columns.

```{r}
D <- D[,
       `:=` (General_Transaction_ID = mask(General_Transaction_ID),
             Physician_Profile_ID = mask(Physician_Profile_ID),
             Physician_Last_Name = mask(Physician_Last_Name))]
```

Show the masked datasets.
With an appropriate hash/anonymizer/scrambler function, the following should be HIPAA-compliant.

```{r}
head(D)
str(D)
```


# Example of writing to an external dataset

Define a output function.

```{r}
outputTo <- function (obj, f, path) {
  outf <- paste(path, f, sep="\\")
  message(sprintf("See output file %s", outf))
  sink(outf)
  show(obj)
  sink()
}
```

Set folder/directory to output data to.
**This folder should be accessible only to individuals covered by the data use agreement (DUA) to use the data.**
For illustration, a temp folder is used, but the code is generalizable.

```{r}
pathProtected <- tempdir()  ## Change this to another protected folder
```

Write `head(D)` to an output file in the protected folder.

```{r}
f <- paste0("exampleOutputHead_", Sys.Date())
outputTo(head(D), f, pathProtected)
```

Write `str(D)` to an output file in the protected folder.

```{r}
f <- paste0("exampleOutputStr_", Sys.Date())
outputTo(str(D), f, pathProtected)
```
